Modeling MOOC Student Behavior With Two-Layer Hidden Markov Models

### 2017-06-22 presentation/presentation.tex
Improve visualization for EDM 2017 presentation.

\begin{frame}{The Back of My Envelope}
  If every course looked like Cheng's\ldots
  \begin{center}
      \[
        \sim 7,000\text{ courses } \times \frac{\sim 85,000\text{ browsing
        sessions }}{\text{ course }} \times \frac{\sim 7\text{
        actions }}{\text{ browsing session }}
      \]
      \Large
      \begin{align*}
        \approx &\text{ } \mathbf{4.2} \text{ \textbf{billion} actions }
        \\
        &\text{ across } \mathbf{595} \text{ \textbf{million} sessions }
      \end{align*}
  \end{center}
\end{frame}

\begin{frame}{Where we come in\ldots}
    We need \textbf{\alert{concise}} and \textbf{\alert{digestible}}
    summaries of these logs in order to understand them!
    \\[\baselineskip]
    This paper contributes the following:
    \begin{enumerate}
      \item A suggestion for a \textbf{behavior pattern representation}, and
      \item A model that can efficiently and \textbf{automatically} capture
          this kind of representation in an \textbf{unsupervised} way
    \end{enumerate}
\end{frame}

@@

\begin{frame}{The Model}
  Notation:
  \begin{itemize}
    \item $\mathbf{A}$: a set of actions students take in the MOOC log,
    \item $\mathbf{L}$: the set of all students in the MOOC,
    \item $\mathbf{O}^{(\ell)}$: the list of action sequences (list of
        lists) for each student $\ell \in \mathbf{L}$
    \item $\mathbf{o} \in \mathbf{O}^{(\ell)}$: a list of actions a student
        took in one ``session'' $\mathbf{o} = (a_1, a_2, \ldots, a_T)$
        where $a_i \in \mathbf{A}$.
  \end{itemize}
\end{frame}

\begin{frame}{The Model}
  Each pattern is associated with a latent state $u \in \{1,\ldots,K\}$,
  which is itself a Markov model represented with $\lambda^{(u)} =
  (\pi^{(u)}, A^{(u)})$ where
  \begin{itemize}
    \item $\pi^{(u)}$ is a $|\mathbf{A}|$ length probability vector to
        indicate the probability of starting with any action $a \in
        \mathbf{A}$, and
    \item $A^{(u)}$ is a $|\mathbf{A}|\times|\mathbf{A}|$ matrix indicating
        the transition probabilities between each pair of actions in
        $\mathbf{A}$.
  \end{itemize}

  \textbf{Assumption}: Each action list $\mathbf{o} \in
  \mathbf{O}^{(\ell)}$ is generated from one specific latent state.
  If $\mathbf{o} = (a_1, \ldots, a_T)$, then
  \[
      P(\mathbf{o} \mid \lambda^{(u)})
      = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^T P(a_i \mid a_{i-1}, A^{(u)})
  \]
\end{frame}

\begin{frame}{The Model}
  Finally, we can compute the likelihood of a list of action sequences
  $\mathbf{O}^{(\ell)}$ of length $N$ for any student $\ell \in \mathbf{L}$
  as
  \begin{align*}
      P(\mathbf{O}^{(\ell)} \mid \Lambda) = \sum_{(v_1,\ldots,v_N) \in U}
          &\left(P(v_1 \mid \Lambda) P(\mathbf{o}_1 \mid
          \lambda^{(v_1)})\phantom{\prod_{i}^N}\right.\\
          & \left.\times
          \prod_{i=2}^N P(v_i \mid v_{i-1}, \Lambda)P(\mathbf{o}_i \mid
      \lambda^{(v_i)})\right)
  \end{align*}
  where $\Lambda = (\pi, A, B)$ is the tuple of all model parameters.
  \begin{itemize}
    \item $\pi$: distribution over initial latent states,
    \item $A$: $K\times K$ transition matrix for each pair of latent
        states,
    \item $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$, the latent state
        Markov model representations
  \end{itemize}
\end{frame}

>>>>

\begin{frame}{The Back of My Envelope}
  Data looks like:
  \begin{center}
    view lecture \rightarrow\, quiz \rightarrow\, forum \rightarrow\,
    \ldots\, \rightarrow\, view lecture \rightarrow\, quiz \rightarrow\,
    quiz \rightarrow\, forum \rightarrow\, \ldots
  \end{center}

  If every course looked like Cheng's\ldots
  \begin{center}
      \[
        \sim 7,000\text{ courses } \times \frac{\sim 85,000\text{ browsing
        sessions }}{\text{ course }} \times \frac{\sim 7\text{
        actions }}{\text{ browsing session }}
      \]
      \Large
      \begin{align*}
        \approx &\text{ } \mathbf{4.2} \text{ \textbf{billion} actions }
        \\
        &\text{ across } \mathbf{595} \text{ \textbf{million} sessions }
      \end{align*}
  \end{center}
\end{frame}

\begin{frame}{Where we come in\ldots}
    We need \textbf{\alert{concise}} and \textbf{\alert{digestible}}
    summaries of these logs in order to understand them!
    \\[\baselineskip]
    This paper contributes the following:
    \begin{enumerate}
      \item A suggestion for a \textbf{behavior pattern representation} to
       understand patterns and \textbf{transitions between them}, and
       \\[0.5\baselineskip]
      \item A model that can efficiently and \textbf{automatically} capture
       this kind of representation in an \textbf{unsupervised} way
    \end{enumerate}
\end{frame}

@@

\begin{frame}{The Generative Model}
  \begin{center}
    \begin{tikzpicture}
      \tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 10mm]
      \tikzstyle{observed}=[circle, minimum size = 10mm, thick, draw=black!80, node distance = 10mm, fill = black!10]
      \tikzstyle{invismain}=[circle, minimum size = 10mm, thick, draw =white!100, node distance = 10mm]
      \tikzstyle{connect}=[-latex, thick]
      \tikzstyle{box}=[rectangle, draw=black!100]

      % tags
      \node[main, fill = white!100] (s1) [] { $s_1$ };
      \node[main, fill = white!100] (s2) [right=of s1] { $s_2$ };
      \node[main, fill = white!100] (s3) [right=of s2] { $s_3$ };
      \node[main, fill = white!100] (s4) [right=of s3] { $s_4$ };
      \node[invismain, opacity=0, text opacity=1] (sn) [right=of s4] {$\ldots$};

      % words
      \node[observed] (o1) [below=of s1] { $\mathbf{o_1}$ };
      \node[observed] (o2) [below=of s2] { $\mathbf{o_2}$ };
      \node[observed] (o3) [below=of s3] { $\mathbf{o_3}$ };
      \node[observed] (o4) [below=of s4] { $\mathbf{o_4}$ };
      \node[invismain, opacity=0, text opacity=1] (on) [below=of sn] {$\ldots$};

      \path (s1) edge [connect] (s2)
        (s2) edge [connect] (s3)
        (s3) edge [connect] (s4)
        (s4) edge [connect] (sn);

      \path (s1) edge [connect] (o1);
      \path (s2) edge [connect] (o2);
      \path (s3) edge [connect] (o3);
      \path (s4) edge [connect] (o4);

    \end{tikzpicture}
  \end{center}
  \[
    \underbrace{\overbrace{\langle \overbrace{lecture}^{\text{action } a_1}
      \rightarrow\, \overbrace{quiz}^{\text{action } a_2} \rightarrow\,
      \ldots\, \rightarrow\, \overbrace{wiki}^{\text{action } a_T}
    \rangle}^{\text{observation } \mathbf{o}_1 \text{ ``generated'' from }
    s_1}, \overbrace{\langle \overbrace{wiki}^{\text{action } a_1}
      \rightarrow\, \overbrace{quiz}^{\text{action } a_2} \rightarrow\,
    \ldots \rangle}^{\text{observation } \mathbf{o}_2 \text{ ``generated'' from } s_2},
    \ldots}_{\text{student sequence } \mathbf{O}^{(\ell)}}
  \]
\end{frame}

\begin{frame}{The Generative Model}
  \begin{center}
    \begin{tikzpicture}
      \tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 10mm]
      \tikzstyle{observed}=[circle, minimum size = 10mm, thick, draw=black!80, node distance = 10mm, fill = black!10]
      \tikzstyle{invismain}=[circle, minimum size = 10mm, thick, draw =white!100, node distance = 10mm]
      \tikzstyle{connect}=[-latex, thick]
      \tikzstyle{box}=[rectangle, draw=black!100]

      % tags
      \node[main, fill = white!100] (s1) [] { $s_1$ };
      \node[main, fill = white!100] (s2) [right=of s1] { $s_2$ };
      \node[main, fill = white!100] (s3) [right=of s2] { $s_3$ };
      \node[main, fill = white!100] (s4) [right=of s3] { $s_4$ };
      \node[invismain, opacity=0, text opacity=1] (sn) [right=of s4] {$\ldots$};

      % words
      \node[observed] (o1) [below=of s1] { $\mathbf{o_1}$ };
      \node[observed] (o2) [below=of s2] { $\mathbf{o_2}$ };
      \node[observed] (o3) [below=of s3] { $\mathbf{o_3}$ };
      \node[observed] (o4) [below=of s4] { $\mathbf{o_4}$ };
      \node[invismain, opacity=0, text opacity=1] (on) [below=of sn] {$\ldots$};

      \path (s1) edge [connect] (s2)
        (s2) edge [connect] (s3)
        (s3) edge [connect] (s4)
        (s4) edge [connect] (sn);

      \path (s1) edge [connect] (o1);
      \path (s2) edge [connect] (o2);
      \path (s3) edge [connect] (o3);
      \path (s4) edge [connect] (o4);

    \end{tikzpicture}
  \end{center}

  Model parameters $\Lambda = (\pi, A, B)$:
  \begin{itemize}
    \item $\pi$: probability of starting student sequence in latent state
      $s_i$
    \item $A$: transition probability between latent states
    \item $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$ where
      \begin{itemize}
        \item $\lambda^{(i)} = (\pi^{(i)}, A^{(i)})$, initial probability
          and transition matrix between student \emph{actions} (e.g. view
          forum, take quiz, \ldots) for latent state $s_i$
      \end{itemize}
  \end{itemize}
\end{frame}
